\textbf{ Mapping stellar surfaces\\ I: Degeneracies in the rotational light curve problem }

### 2021-01-25 paper1/ms.tex
more text

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=0.65\linewidth]{figures/rank.pdf}
        \oscaption{rank}{%
            Rank and nullity of the flux operator. The orange curve
            shows the number of spherical harmonic coefficients required
            to fully describe a stellar surface up to a given spherical
            harmonic degree (bottom axis) or, equivalently, up to an
            effective surface resolution (top axis). The blue curve shows
            the rank of the flux operator, corresponding to the maximum
            number of independent signals that can be constrained from a
            light curve. The size of the null space (the nullity) is
            the difference between the two curves.
            \label{fig:rank}
        }
    \end{centering}
\end{figure}

@@

@@

@@

@@

@@

\begin{equation}

@@

\end{equation}

@@

So what does this all mean? By performing SVD, we separated
the surface map representation $\mathbf{y}$ into orthogonal components
$\mathbf{y}_\bullet$ and $\mathbf{y}_\circ$, consisting of the
terms that contribute to the light curve $\mathbf{f}$ and those
that don't, respectively. In particular, it is useful to consider
the linear operators we used to perform this decomposition
in Equations~(\ref{eq:yrow}) and (\ref{eq:ynull}):
%
\begin{align}
    \mathbf{P} & \equiv \mathbf{V}_\bullet \mathbf{V}_\bullet^\top
    \\
    \mathbf{N} & \equiv \mathbf{V}_\circ \mathbf{V}_\circ^\top
    \quad,
\end{align}
%
which we will call the \emph{preimage operator} and \emph{null space
    operator}, respectively. The preimage operator $\mathbf{P}$
transforms a vector $\mathbf{y}$ in the surface map basis in such a way
that it preserves the information in $\mathbf{y}$ that gets mapped
onto the light curve $\mathbf{f}$ via $\mathbf{A}$ (the \emph{preimage}) and discards the rest. The
null space operator $\mathbf{N}$ does the opposite: it preserves only
the information in $\mathbf{y}$ that gets mapped onto the zero
vector via $\mathbf{A}$ (the \emph{null space}).
%
In other words,
the $\mathbf{P}$ and $\mathbf{N}$ operators reveal the
components of the surface map $\mathbf{y}_\bullet$ that contribute to the
light curve and the components $\mathbf{y}_\circ$ that don't.
Likewise, $\mathbf{y}_\bullet$ represents all the information that can be
learned from a stellar light curve, and $\mathbf{y}_\circ$ represents all the
information that cannot.

@@

\subsection{Implications for inference}
%
It is useful to consider Figure~\ref{fig:nullspace_preimage} in the
context of an inference exercise. One can equivalently think of
the preimage as the solution to a least squares inference problem
in which a completely uninformative prior%
\footnote{%
    More specifically, a zero-mean, infinite variance Gaussian prior on the coefficients of $\mathbf{y}$.
}
is placed on the spherical harmonic
coefficients. In this case, the least squares solution is given by
%
\begin{proof}{test_svd}
    \label{eq:inference}
    \hat{\mathbf{y}}
    &=
    \lim_{\lambda \rightarrow 0} \left( \mathbf{A}^\top \mathbf{A} + \lambda \mathbf{I}\right)^{-1} \mathbf{A}^\top \mathbf{f}
    \nonumber\\
    &= \mathbf{y}_\bullet
    \quad.
\end{proof}
%
In this sense, $\mathbf{y}_\bullet$ represents our knowledge about the
surface of a star after an observation if we have no prior information
whatsoever on $\mathbf{y}$.
\emph{Any other information about the surface must come from the prior.}%

Equation~(\ref{eq:inference}) is the solution to the surface map
we would obtain if we knew \emph{nothing} about the stellar surface
before analyzing the light curve. In practice, this is hardly,
if ever, the case. For instance, we know that stellar surfaces
must have non-negative intensities everywhere. While this may seem
like a trivial prior, non-negativity can be a powerful degeneracy-breaking
constraint \citep[e.g.,][]{Fienup1982}. Moreover, we know that
stellar surfaces usually consist of localized features; under an
appropriate compactness prior, solutions
like the preimage in the bottom row of
Figure~\ref{fig:nullspace_preimage} (for example) could be
confidently ruled out.

@@

Instead of lamenting the difficulties of constraining stellar surface features
at near-edge-on orientations, let us focus on the fact that the
null space is a function of inclination.
%
A useful property of
the spherical harmonics is that rotation about any axis can only change the
order $m$ of the harmonic, while keeping the degree $l$ fixed: in other
words, rotation simply mixes the spherical harmonics along rows in
Figure~\ref{fig:ylms}. Therefore, if we rotate the spherical harmonics
in the figure about a different axis, we are still subject to just as
many degeneracies (i.e., the size of the null space is unchanged),
but the particular linear combination of modes in the null space will
be different.

\xxx{more...}

Let us define the \emph{posterior shrinkage} $S$ as
%
\begin{align}
    \label{eq:shrinkage}
    S \equiv 1 - \lim\limits_{\sigma_0^2 \rightarrow \infty}
    \frac{\sigma^2}{\sigma_0^2}
    \quad,
\end{align}
%
where $\sigma_0^2$ is the prior variance
and $\sigma^2$ is the posterior variance \citep[see, e.g.,][]{Betancourt2018}.
The posterior shrinkage is a measure of how informative a measurement is about a
given mode on the surface in the limit of infinite signal-to-noise ratio (SNR)
and is independent of what the stellar surface actually looks like.
If, at infinite SNR and with a completely uninformative prior,
a particular mode can be learned exactly from a dataset, the posterior
shrinkage is defined to be unity. Conversely, if the data is completely
unconstraining of that mode (i.e., it is entirely in the null space),
$S$ will tend to zero.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_single.pdf}
        \oscaption{nullspace_ensemble_single}{%
            Posterior shrinkage...
            \label{fig:nullspace_ensemble_single}
        }
    \end{centering}
\end{figure}

\xxx{Discuss Figure~\ref{fig:nullspace_ensemble_single}, and how it shows
    the modes we can learn and those we can't. Can't learn odd modes, can't
    usually learn much about high degree stuff. Variance is due to different
    inclinations. What if we could harness that?}

This all seems quite hopeless, and for analyses of individual stars,
it pretty much is. Unless extremely strong (and likely incorrect) priors are employed
(such as assuming the stellar surface consists of exactly two perfectly
circular, uniform contrast spots), we simply cannot learn much about
the surface of an individual star from photometry.

However, we can use the dependence of the null space on inclination to our
advantage. If we could observe a star from many different vantage points,
we would be able to break many of the degeneracies at play, since we
would get different constraints on the amplitude of each mode when viewed
at different inclinations. This, of course, is not possible (at least not
yet!). But what we \emph{can} do is observe many similar stars, each viewed
at a different (random) inclination, and attempt to learn something about
the properties of the ensemble of stars as a whole.
In the
following section, we explore the role of ensemble analyses in breaking
the degeneracies of the mapping problem in more detail.

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure the light curves
            of 1 (blue), 3 (orange), 10 (green), and 30 (red) \emph{identical} stars
            viewed at random orientations.
            The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing many identical stars from different vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes. This is why ensemble analyses of
            many similar stars at different inclinations allows us to infer
            their surface properties.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

Figure~\ref{fig:nullspace_ensemble} shows the posterior shrinkage
as a function of spherical harmonic degree for a few different hypothetical
scenarios in which we observe one or more stellar light curves with
no limb darkening.
Let us first consider the blue curves, in which a light curve is
collected from a single star viewed at a single random inclination (which, of course, is
all we can do from our fixed vantage point on Earth).
Each thin blue curve corresponds to a particular random draw from the (isotropic)
inclination distribution; the thick blue curve is the average over 300 trials.
For some of the low-degree modes, the shrinkage is relatively high: it is
fairly easy to constrain the dipole moment from a light curve, as this is
usually the dominant sinusoidal signal. However, as the degree $l$
increases, the shrinkage decreases dramatically: at $l = 14$, corresponding
to features on scales of roughly $15^\circ$, the light curve
can only tell us about $\sim 10\%$ of the total information about what the
surface looks like. As $l$ increases further, $S$ tends to zero.
Another important feature of the shrinkage, which we hinted at above,
is that it is exactly zero for
all odd-degree modes above $l = 1$. This is a well-known fact: all odd spherical
harmonics other than the dipole are in the null space \emph{regardless of
    inclination} \citep[e.g.,][]{Luger2019}. In other words, these spherical
harmonics are perfectly antisymmetric in projection over the unit disk
when viewed from any orientation. Absent structure to break these symmetries
(see below), we simply cannot learn anything about these modes from
stellar light curves. If we average over the shrinkage for all modes up to $l=15$,
we find that a single light curve measurement can only tell us $\sim 9\%$
of the information about the surface on those scales.

In an ensemble analysis, we assume we observe the lightcurves of many stars
that are ``similar'' in some statistical sense. As a thought experiment,
let us consider an extreme version of ensemble analysis in which all the
stars in our dataset happen to have \emph{identical} surfaces. We will
still assume they are oriented at random inclinations, as we would expect for
field stars. The orange curves in the top panel of the figure correspond to
the case in which we analyze ensembles consisting of three identical stars
viewed at three different random inclinations.
Interestingly, the shrinkage increased at all even spherical harmonic degrees
(the odd degrees, as we mentioned above, are always invisible).
Note that since we are in the limit of infinite SNR, the fact that we have
three times the data as we did in the previous scenario is irrelevant: the
increase in the shrinkage is instead due to the fact that our observations
from different vantage points broke some degeneracies in the problem.
This is a consequence of the fact we mentioned in the previous section:
the null space (for the even modes)
is a strong function of the inclination.
If we average over all modes, we obtain a total shrinkage of $\sim 24\%$
for $l\leq15$ in this case.

Finally, the green and red curves correspond respectively to ensembles of ten and thirty
identical stars observed at random inclinations.
In these scenarios, the posterior shrinkage
is nearly $100\%$
for all even modes (it asymptotically approaches $100\%$ as the number
of light curves increases further). The total shrinkage is $\sim 47\%$
and approaches $50\%$ as both $l_\mathrm{max}$  and the number of light curves
tend to infinity. Thus, if we were able to measure light curves of identical stars
from many different inclinations, the null space would consist \emph{only}
of the odd modes, and our data would tell us exactly half of all the
information about the surfaces of these objects.

Figure~\ref{fig:nullspace_preimage_ensemble} shows \xxx{...}
Because spots are compact features, they
are necessarily made up of a continuum of spherical harmonic modes spanning
many different values of $l$: they can therefore be seen in both the even
modes (the preimage) and the odd modes (the null space). Interestingly,
the symmetries at play require spots to be paired with antipodal dark mirror
images in the even modes, and with \emph{bright} ones in the odd modes
(which sum to perfectly cancel out in the true map). Since these reflections
are at the same latitude (in the opposite hemisphere) as the true spots
\xxx{...}

@@

\begin{figure}[p!]
    \begin{centering}
        %
        \textbf{Ensemble of light curves at many inclinations}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_a.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_b.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_c.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_d.pdf}
        \oscaption{nullspace_preimage_ensemble}{%
            Same as Figure~\ref{fig:nullspace_preimage}, \xxx{...}
            \label{fig:nullspace_preimage_ensemble}
        }
    \end{centering}
\end{figure}

\subsection{Limb darkening}
\label{sec:limbdark}

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld_var.pdf}
        \oscaption{nullspace_ensemble_ld_var}{%
            Same as Figure~\ref{fig:nullspace_ensemble_ld}, but allowing for
            a 10\% variation in $u_1$ and $u_2$
            across different observations. For $\gtrsim 30$ light curves,
            there is virtually no null space up to at least $l_\mathrm{max} = 15$.
            \label{fig:nullspace_ensemble_ld_var}
        }
    \end{centering}
\end{figure}

As we saw in Figure~\ref{fig:calibration_ld}, the presence of limb darkening
makes it significantly more difficult to infer spot properties at fixed
SNR. To understand
why, let us repeat the experiment from the previous section,
this time with moderate quadratic limb darkening ($u_1 = 0.5$ and $u_2 = 0.25$).
The top panel of Figure~\ref{fig:nullspace_ensemble_ld} shows the
posterior shrinkage plot (same as Figure~\ref{fig:nullspace_ensemble}, but
with limb darkening). Interestingly, there is no longer a clean division of
the null space between even and odd modes in the limit of large $M$.
This is because limb darkening effectively lifts odd modes out of the null space, \emph{at the
    expense of the even modes}. While no coefficient lies entirely in the
null space ($R = 0$) when limb darkening is present, no coefficient
can be uniquely inferred ($R = 1$), either. This can be understood by noting that
a polynomial limb darkening law can be written
exactly as a linear combination of the $m=0$ spherical harmonics
up to a degree equal to the order of the limb darkening
\citep[in this case, $l = 2$;][and Appendix~\ref{sec:ld}]{Luger2019,Agol2020}.
Since the limb darkening operation is a (multiplicative) downweighting of
the surface intensity, the map seen by the observer is just the product
of the spherical harmonic representation of the surface ($\mathbf{f}$)
and the spherical harmonic representation of the limb darkening profile.
And since spherical harmonics are just polynomials on the surface of the sphere,
the product of spherical harmonics of degree $l_1$ and $l_2$ is a spherical
harmonic of degree $l_1 + l_2$. This means that the linear limb darkening
component ($l = 1$) effectively raises the degree of all spherical harmonic
coefficients of the surface map by one. This has the effect of reversing
the null space: under \emph{only} linear limb darkening, it is the \emph{even}
modes that would like in the null space. However, the quadratic
limb darkening term ($l = 2$) raises the degree of all spherical harmonics by two,
so its presence ensures that the even modes can still be probed to some extent.
In reality, the true limb darkening profile of a stellar surface is more
complicated than a two-parameter quadratic model can capture; but one may still
expand it as an arbitrary order polynomial, in which case the argument still
applies---limb darkening mixes the null space and the preimage in a nontrivial
way. The fact that no coefficient can be determined uniquely---i.e., there are
perfect degeneracies involving \emph{all} modes on the surface---is what makes it more
difficult in practice to perform ensemble analyses on limb-darkened stars.
Nevertheless, it is still possible to uniquely and precisely infer the spot properties
under limb darkening,
but as a rule of thumb, the stronger the limb darkening, the larger the
ensemble must be (see Figure~\ref{fig:calibration_ld_1000}).

In reality, it is unlikely that all stars in a given ensemble have exactly
the same limb darkening coefficients, however ``similar'' the stars may be.
The bottom panel of Figure~\ref{fig:nullspace_ensemble_ld} shows the same
posterior shrinkage plot, but for the case where each star has coefficients
$u_1 = 0.5 \pm 0.05$ and $u_2 = 0.25 \pm 0.025$; i.e., we add a scatter of
10\% in the value of these coefficients. The plot shows the posterior
shrinkage in the hypothetical case where we know the coefficients for
each star exactly. Now, as the size of the ensemble increases, the posterior
shrinkage approaches unity \emph{for all spherical harmonic modes}.
In the same way that the null space is a strong function of the inclination,
allowing us to chip away at it with observations at different inclinations,
the null space is also a strong function of the limb darkening law. Even a
small amount of variance in the coefficients is sufficient to constrain all surface
modes exactly (in the limit of infinite SNR and large $M$).

In practice, of course, we will never know the limb darkening coefficients
exactly, so we must either model them or marginalize over them
(see \S\ref{sec:other-marg}). This will dilute our ability to infer surface
properties exactly.

@@

>>>>

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/rank.pdf}
        \oscaption{rank}{%
            Rank and nullity of the flux operator. The orange curve
            shows the number of spherical harmonic coefficients required
            to fully describe a stellar surface up to a given spherical
            harmonic degree (bottom axis) or, equivalently, up to an
            effective surface resolution (top axis). The blue curve shows
            the rank of the flux operator, corresponding to the maximum
            number of independent signals that can be constrained from a
            light curve. The size of the null space (the nullity) is
            the difference between the two curves.
            \label{fig:rank}
        }
    \end{centering}
\end{figure}

@@

\begin{proof}{test_decomposition}

@@

\end{proof}

@@

\begin{proof}{test_PN}

@@

\begin{proof}{test_orthogonality}

@@

\begin{proof}{test_decomposition}

@@

\end{proof}

@@

So what does this all mean? By performing SVD, we separated
the surface map representation $\mathbf{y}$ into orthogonal components
$\mathbf{y}_\bullet$ and $\mathbf{y}_\circ$, consisting of the
terms that contribute to the light curve $\mathbf{f}$ and those
that don't, respectively. In particular, it is useful to consider
the linear operators we used to perform this decomposition
in Equations~(\ref{eq:yrow}) and (\ref{eq:ynull}):
%
\begin{proof}{test_PN}
    \mathbf{P} & \equiv \mathbf{V}_\bullet \mathbf{V}_\bullet^\top
    \\
    \mathbf{N} & \equiv \mathbf{V}_\circ \mathbf{V}_\circ^\top
    \quad,
\end{proof}
%
which we will call the \emph{preimage operator} and \emph{null space
    operator}, respectively. The preimage operator $\mathbf{P}$
transforms a vector $\mathbf{y}$ in the surface map basis in such a way
that it preserves the information in $\mathbf{y}$ that gets mapped
onto the light curve $\mathbf{f}$ via $\mathbf{A}$ (the \emph{preimage}) and discards the rest. The
null space operator $\mathbf{N}$ does the opposite: it preserves only
the information in $\mathbf{y}$ that gets mapped onto the zero
vector via $\mathbf{A}$ (the \emph{null space}).
%
In other words,
the $\mathbf{P}$ and $\mathbf{N}$ operators reveal the
components of the surface map $\mathbf{y}_\bullet$ that contribute to the
light curve and the components $\mathbf{y}_\circ$ that don't.
Likewise, $\mathbf{y}_\bullet$ represents all the information that can be
learned from a stellar light curve, and $\mathbf{y}_\circ$ represents all the
information that cannot.

@@

\subsection{Implications for inference}
%
It is useful to consider Figure~\ref{fig:nullspace_preimage} in the
context of an inference exercise. One can equivalently think of
the preimage as the solution to a least squares inference problem
in which a completely uninformative prior%
\footnote{%
    More specifically, a zero-mean, infinite variance Gaussian prior on the coefficients of $\mathbf{y}$.
}
is placed on the spherical harmonic
coefficients. In this case, the least squares solution is given by
%
\begin{proof}{test_svd_lstsq}
    \label{eq:inference}
    \hat{\mathbf{y}}
    &=
    \lim_{\lambda \rightarrow 0} \left( \mathbf{A}^\top \mathbf{A} + \lambda \mathbf{I}\right)^{-1} \mathbf{A}^\top \mathbf{f}
    \nonumber\\
    &= \mathbf{y}_\bullet
    \quad.
\end{proof}
%
In this sense, $\mathbf{y}_\bullet$ represents our knowledge about the
surface of a star after an observation if we have no prior information
whatsoever on $\mathbf{y}$.
\emph{Any other information about the surface must come from the prior.}%

Equation~(\ref{eq:inference}) is the solution to the surface map
we would obtain if we knew \emph{nothing} about the stellar surface
before analyzing the light curve. In practice, this is never really
the case. For instance, we know that stellar surfaces
must have non-negative intensities everywhere. While this may seem
like a trivial prior, non-negativity can be a powerful degeneracy-breaking
constraint \citep[e.g.,][]{Fienup1982}. Moreover, we know that
stellar surfaces usually consist of localized features; under an
appropriate compactness prior, solutions
like the preimage in the bottom row of
Figure~\ref{fig:nullspace_preimage} (for example) could be
confidently ruled out.

@@

Instead of lamenting the difficulties of constraining stellar surface features
at near-edge-on orientations, let us focus on the fact that the
null space is a function of inclination.
%
A useful property of
the spherical harmonics is that a rotation operation on any
component in the basis can only change the order $m$ of the
harmonic; the degree $l$ is constant under rotation. In other
words, rotating any of the spherical harmonics in Figure~\ref{fig:ylms}
about an arbitrary axis simply yields a weighted sum of the spherical
harmonics along its row.%
\footnote{For instance, rotation of $Y_{1,-1}$ by $90^\circ$ about
    $\hat{\mathbf{x}}$ yields $Y_{1,1}$; other rotations will in general
    yield a weighted combination of $Y_{1,-1}$, $Y_{1,0}$, and $Y_{1,1}$.}
This means that changing the inclination of the star---which changes
the axis of rotation in the observer's frame---simply changes the weighting
of modes that give rise to certain signals in the light curve.
This, in turn, results in the dependence of the null space on inclination.

To better understand this effect, let us
define the \emph{posterior shrinkage} $S$ as
%
\begin{align}
    \label{eq:shrinkage}
    S \equiv 1 - \lim\limits_{\sigma_0^2 \rightarrow \infty}
    \frac{\sigma^2}{\sigma_0^2}
    \quad,
\end{align}
%
where $\sigma_0^2$ is the prior variance
and $\sigma^2$ is the posterior variance
on a particular surface mode we are trying to constrain
\citep[see, e.g.,][]{Betancourt2018}.
The posterior shrinkage is a measure of how informative a measurement is about a
given mode on the surface in the limit of infinite signal-to-noise ratio (SNR)
and is independent of what the stellar surface actually looks like.
If, at infinite SNR and with a completely uninformative prior,
a particular mode can be learned exactly from a dataset, the posterior
shrinkage is defined to be unity. Conversely, if the data is completely
unconstraining of that mode (i.e., it is entirely in the null space),
$S$ will tend to zero.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_single.pdf}
        \oscaption{nullspace_ensemble_single}{%
            Posterior shrinkage as a function of spherical harmonic
            degree $l$ given a single observation of a star at a random
            inclination (thin blue curves). The mean shrinkage is shown
            as the thicker curve. The information content of light curves
            tends to zero as $l$ increases, and odd $l > 1$ modes are in the
            null space at all inclinations.
            \label{fig:nullspace_ensemble_single}
        }
    \end{centering}
\end{figure}

Figure~\ref{fig:nullspace_ensemble_single} shows the posterior shrinkage
given a single observation of a star at a random inclination.
Each thin blue curve corresponds to a particular draw from an isotropic
inclination distribution; the thick blue curve is the average over 300 trials.
For some of the low-degree modes, the shrinkage is relatively high: it is
fairly easy to constrain the dipole moment from a light curve, as this is
usually the dominant sinusoidal signal. However, as the degree $l$
increases, the shrinkage decreases dramatically: at $l = 14$, corresponding
to features on scales of roughly $13^\circ$, the light curve
can only tell us about $\sim 10\%$ of the total information about what the
surface looks like. As $l$ increases further, $S$ tends to zero.
Another important feature of the shrinkage, which we hinted at above,
is that it is exactly zero for
all odd-degree modes above $l = 1$. This is a well-known fact: all odd spherical
harmonics other than the dipole are in the null space \emph{regardless of
    inclination} \citep[e.g.,][]{Luger2019}. In other words, these spherical
harmonics are perfectly antisymmetric in projection over the unit disk
when viewed from any orientation. Absent structure to break these symmetries
(see below), we simply cannot learn anything about these modes from
stellar light curves. If we average over the shrinkage for all modes up to $l=15$,
we find that a single light curve measurement can only tell us $\sim 9\%$
of the information about the surface on those scales.

Fortunately, however, there is quite a bit of variance in the shrinkage
for different values of the stellar inclination.
As we will see, we can use this dependence of the null space on inclination to our
advantage. If we could observe a star from many different vantage points,
we would be able to break many of the degeneracies at play, since we
would get different constraints on the amplitude of each mode when viewed
at different inclinations. This, of course, is not possible (at least not
yet!). But what we \emph{can} do is observe many similar stars, each viewed
at a different (random) inclination, and attempt to learn something about
the properties of the ensemble of stars as a whole.
In the
following section, we explore the role of ensemble analyses in breaking
the degeneracies of the mapping problem in more detail.

@@

In an ensemble analysis, we assume we observe the lightcurves of many stars
that are ``similar'' in some statistical sense. As a thought experiment,
let us consider an extreme version of ensemble analysis in which all the
stars in our sample happen to have \emph{identical} surfaces. We will
still assume they are oriented at random inclinations, as we would expect for
field stars.
%
Figure~\ref{fig:nullspace_ensemble} shows the posterior shrinkage
curves for this hypothetical scenario, assuming we have access to
light curves of 1 (blue), 3 (orange), 10 (green) and 30 (red)
identical stars viewed at random inclinations.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Similar to Figure~\ref{fig:nullspace_ensemble_single}, but
            assuming the observer can measure the light curves
            of 1 (blue), 3 (orange), 10 (green), and 30 (red) \emph{identical} stars
            viewed at random orientations.
            As we saw in Figure~\ref{fig:nullspace_ensemble_single},
            the information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing many identical stars from different vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes. This is why ensemble analyses of
            many similar stars at different inclinations allows us to infer
            their surface properties.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

The addition of light curve measurements at different orientations
increases the posterior shrinkage at all even spherical harmonic degrees
(the odd degrees, as we mentioned above, are always invisible).
Note that since we are in the limit of infinite SNR, the fact that we have
more light curves (i.e., more data) is irrelevant: the
increase in the shrinkage is instead due to the fact that our observations
from different vantage points broke some degeneracies in the problem.
This is a consequence of the fact we mentioned in the previous section:
the null space (for the even modes)
is a strong function of the inclination.

If we average over all modes, we obtain a total shrinkage of $\sim 24\%$
for $l\leq15$ when our sample size is 3 (orange curves): we have more than doubled the
information content of our observations. If we further increase our sample size
to 10 (green curves), the posterior shrinkage approaches $100\%$ for all even $l\leq15$
modes, effectively saturating for a sample size of 30 (red curves).
Thus, if we were able to measure light curves of identical stars
from many different inclinations, the null space would consist \emph{only}
of the odd modes. In the limit of a large number of light curves,
and assuming all stars in the sample have identical surfaces,
\textbf{ensemble analyses can tell us up to
    half of all the information about the stellar surfaces.}

\begin{figure}[p!]
    \begin{centering}
        %
        \textbf{Ensemble of light curves at many inclinations}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_a.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_b.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_c.pdf}
        \\[1em]
        \includegraphics[width=\linewidth]{figures/nullspace_preimage_d.pdf}
        \oscaption{nullspace_preimage_ensemble}{%
            Same as Figures~\ref{fig:nullspace_preimage} and \ref{fig:nullspace_preimage_85}, \
            but assuming we can measure the light curves of these stars from many
            different inclinations. In this limit, the information content of our
            data approaches $50\%$ of the spatial information about the surface.
            \label{fig:nullspace_preimage_ensemble}
        }
    \end{centering}
\end{figure}

To understand what we can learn about the surfaces in this limit, let us
return to the stellar surfaces we considered in Figures~\ref{fig:nullspace_preimage}
and \ref{fig:nullspace_preimage_85}.
Figure~\ref{fig:nullspace_preimage_ensemble} shows the same decomposition of
these surfaces into preimage and null space, but this time assuming we measure
the light curves of these stars from many different random inclinations.
As expected, the partition of information between the preimage and the nullspace
is about $50{-}50$.
Because spots are compact features, they
are necessarily made up of a continuum of spherical harmonic modes spanning
many different values of $l$: they can therefore be seen in both the even
modes (the preimage) and the odd modes (the null space). The absence of
information about the odd modes therefore does not affect our ability to
infer the shape and location of the features on the surface.%
\footnote{
    There is even some hope of deciphering complex alien messages
    (last row in the figure) in this limit!
}
Interestingly, however,
the symmetries at play require spots to be paired with antipodal dark mirror
images in the preimage, and with \emph{bright} ones in the null space
(which sum to perfectly cancel out in the true map). Thus, there are still
degeneracies in the ensemble problem, but they are much easier to break
with a suitable choice of prior. For instance, in the single spot case
(top row), the ``ghost'' image in the southern hemisphere is surrounded by
a bright ring (whose effect is to cancel out its contribution to the flux);
either a compactness prior or a prior that enforces uniformity in the
background could easily penalize that feature in the fit. This may be
much harder to do for surfaces like that shown in the second row, but
we can still (in principle) learn about the size, shape, and latitude
(if not the number) of starspots from the light curves.

There are several practical reasons why this kind of ensemble analysis
may be very difficult (but not impossible!) in practice. Before
we address those, let us consider an important effect that we have not
addressed thus far in this paper: limb darkening.

\subsection{The effect of limb darkening}
\label{sec:limbdark}

Typically, the shallower the angle between the line of sight
and the stellar surface normal, the higher up in the stellar
atmosphere optical depth unity is reached. At optical wavelengths,
lines of sight directed toward the limb of the star therefore
probe cooler temperatures, resulting in the well-known effect of
\emph{limb darkening}. Features close to the limb of the star therefore
contribute less to the total outgoing flux, and this must be
accounted for when computing the effect of a rotating starspot
on the light curve. Limb darkening is often parametrized
as a low-order polynomial in the cosine of the line of sight angle
\citep{Kopal1950}.

To understand how limb darkening affects our ability to infer
surface properties from stellar light curves,
let us repeat the experiment from the previous section,
this time with moderate quadratic limb darkening (with coefficients
$u_1 = 0.5$ and $u_2 = 0.25$).
The top panel of Figure~\ref{fig:nullspace_ensemble_ld} shows the
posterior shrinkage plot (same as Figure~\ref{fig:nullspace_ensemble}, but
this time accounting for limb darkening). Interestingly, there is no longer a clean division of
the null space between even and odd modes in the limit of a large number
of light curves.
This is because limb darkening effectively lifts odd modes out of the null space, \emph{at the
    expense of the even modes}. While no coefficient lies entirely in the
null space ($S = 0$) when limb darkening is present, no coefficient
can be uniquely inferred ($S = 1$), either. This can be understood by noting that
a polynomial limb darkening law can be written
exactly as a linear combination of the $m=0$ spherical harmonics
up to a degree equal to the order of the limb darkening
\citep[in this case, $l = 2$;]{Luger2019,Agol2020}.
Since the limb darkening operation is a (multiplicative) downweighting of
the surface intensity, the map seen by the observer is just the product
of the spherical harmonic representation of the surface ($\mathbf{y}$)
and the spherical harmonic representation of the limb darkening profile.
And since spherical harmonics are just polynomials on the surface of the sphere,
the product of spherical harmonics of degree $l_1$ and $l_2$ is a spherical
harmonic of degree $l_1 + l_2$. This means that the linear limb darkening
component ($l = 1$) effectively raises the degree of all spherical harmonic
coefficients of the surface map by one. This has the effect of reversing
the null space: under \emph{only} linear limb darkening, it is the \emph{even}
modes that would be in the null space. However, the quadratic
limb darkening term ($l = 2$) raises the degree of all spherical harmonics by two,
so its presence ensures that the even modes can still be probed to some extent.
In reality, the true limb darkening profile of a stellar surface is more
complicated than a two-parameter quadratic model can capture; but one may still
expand it as an arbitrary order polynomial, in which case the argument still
applies---limb darkening mixes the null space and the preimage in a nontrivial
way.
%
The fact that no coefficient can be determined uniquely---i.e., there are
perfect degeneracies involving \emph{all} modes on the surface---could make it more
difficult in practice to perform ensemble analyses on limb-darkened stars.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld_var.pdf}
        \oscaption{nullspace_ensemble_ld_var}{%
            Same as Figure~\ref{fig:nullspace_ensemble_ld}, but allowing for
            a 10\% variation in the limb darkening coefficients $u_1$ and $u_2$
            across different observations. For $\gtrsim 30$ light curves,
            there is virtually no null space up to at least $l_\mathrm{max} = 15$.
            \label{fig:nullspace_ensemble_ld_var}
        }
    \end{centering}
\end{figure}

In reality, it is unlikely that all stars in a given ensemble will have exactly
the same limb darkening coefficients, however ``similar'' the stars may be.
Figure~\ref{fig:nullspace_ensemble_ld_var} shows the same
posterior shrinkage plot as Figure~\ref{fig:nullspace_ensemble_ld},
but for the case where each star has coefficients
$u_1 = 0.5 \pm 0.05$ and $u_2 = 0.25 \pm 0.025$; i.e., we add a scatter of
10\% in the value of these coefficients. The plot shows the posterior
shrinkage in the hypothetical case where we know the coefficients for
each star exactly. Now, as the size of the ensemble increases, the posterior
shrinkage approaches unity \emph{for all spherical harmonic modes}.
In the same way that the null space is a strong function of the inclination,
allowing us to chip away at it with observations at different inclinations,
the null space is also a strong function of the limb darkening law. Even a
small amount of variance in the coefficients is sufficient to constrain all surface
modes exactly (in the limit of infinite SNR and a large number of light curves).

In practice, of course, we will never know the limb darkening coefficients
exactly. In the next section, we will revisit this and other assumptions we
made above in a more sober light.

\subsection{A reality check}
\label{sec:reality-check}

There are three major points that make the kind of ensemble
analyses discussed above difficult in practice. First, and perhaps
most obviously, stars are
not identical, no matter how ``similar'' we think they may be.
Even stars of the same spectral type, age, and metallicity will
in general have different configurations of spots on their surfaces.

@@

Also, we don't know the inclinations.

@@

Nevertheless, it is still possible to uniquely and precisely infer the spot properties
under limb darkening,
but as a rule of thumb, the stronger the limb darkening, the larger the
ensemble must be.
